An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-21 paper.tex
First version of Characterising Participant Interactions, from IMC 2021, plus structure outline for later parts

%..................................................................................................
\pb{Arrival of new authors:}
Finally, Figure~\ref{fig:author_new} shows the percentage of authors in each
year that have not previously authored an RFC. Given that the dataset used here
begins in 2001, 100\% of authors are new in that year. The more stable trend in
recent years likely highlights the churn in RFC authorship, with around 30\%
of authors each year having never previously authored an RFC.

@@

>>>>

%..................................................................................................
% IMC 2021 paper Section 3.2 (last part)
\pb{Arrival of new authors:}
Finally, Figure~\ref{fig:author_new} shows the percentage of authors in each
year that have not previously authored an RFC. Given that the dataset used here
begins in 2001, 100\% of authors are new in that year. The more stable trend in
recent years likely highlights the churn in RFC authorship, with around 30\%
of authors each year having never previously authored an RFC.
\todo{Not clear this fits here}

@@

%==================================================================================================
\section{Organisational Dynamics of the IETF}

%--------------------------------------------------------------------------------------------------
\subsection{Characterising Participant Interactions}

% IMC 2021 paper Section 3.3

The operations of the IETF is largely underpinned by mailing list
interactions, which are used to discuss and finalise the drafts that
eventually become RFCs. Our data, which starts from 1995, confirms their
vital role, with 1,153 mailing lists, containing 2,439,240 emails from
74,646 unique email addresses.

\pb{Volume of Discussion:}
Figure~\ref{fig:pid_count_emailing_yearly} presents the number of emails
sent across the last 25 years, showing that email volumes have grown
significantly with time, plateauing at around 130,000 annual messages since
around 2010. The figure also shows how the number of unique Person ID
counts (as described in \S\ref{sec:data}) observed every year in the mail
lists, showing a decreasing trend in the number of contributors.

Figure~\ref{fig:emailvol_by_year_catg} shows the relative email volume by
type. We see that an increasing fraction of messages originate from
automated addresses. This is largely as a result of the growing use of
GitHub, and other version control systems, for managing drafts. There were
122 active IETF working groups at the time of writing: of these, 17 listed
a GitHub repository in their metadata. The QUIC working group, as one
example, has replaced the typical email list discussions with GitHub
issues: indeed, this is a significant part of the surge observed in 2016.
Given that most working groups use mailing lists to manage their activity,
we do not further analyse interactions that take place on GitHub in this
paper. As a result, Figure~\ref{fig:emailvol_by_year_catg} understates the
volume of interactions: the plateau observed in recent years is at least
somewhat attributable to the shift to GitHub and similar services. As this
shift continues, it will become important for future work to consider these
interactions.
\todo{We have more discussion of GitHub?}

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/pid_count_emailing_yearly.pdf}
\caption{Number of Person IDs exchanging emails each year}
\label{fig:pid_count_emailing_yearly}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/frequency_emails_yearly_categories2.pdf}
\caption{Number of messages exchanged per year across different categories:
  messages mapped to the Datatracker (Datatracker Person-ID); messages by
  automated email addresses (Automated); messages by role-based addresses
  (Role-based); and messages not mapped to the Datatracker (New Person-ID)}
\label{fig:emailvol_by_year_catg}
\end{figure}

%..................................................................................................
\pb{Discussion of drafts:}
The mailing lists are, other than plenary or interim meetings, the primary
means for the discussion of drafts.  To measure how often drafts are
discussed, we next identify mentions of drafts in mailing list messages.
We therefore extract any mention of a draft (beginning \texttt{draft-}) or
RFC (i.e., ``RFC'' followed by a number).  Figure~\ref{fig:draft_mentions}
presents the number of drafts mentions in the emails per year. Separate
mentions of the same draft are counted as different mentions, as we want to
observe the entire volume of mentions.  We observe a strong increase in the
number of mentions over time.  This is largely driven by the growing number
of drafts being published. In fact, we find a Pearson correlation of 0.89
between the number of drafts published and the number of mentions. This
speaks to the influence of emails that mention drafts in driving draft
production.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/yearly-draft-mention-volume.pdf}
\caption{Number of draft mentions in each year found in the mail lists}
\label{fig:draft_mentions}
\end{figure}

%..................................................................................................
\pb{Contribution duration.}
We now look into the longevity of contributors, in terms of the number of
years that they actively participate in any IETF mailing lists. We define
the \textit{contribution duration} of a participant as the length of time
that they have contributed to the mailing lists.  To do this, we look at
the contributors who first contribute to mailing lists between the years
2000 to 2013. We limit our analysis to 2013, since the longevity of
contributors that first participate more recently cannot be determined.
For each year, in the period 2000 to 2013, we look at those who first
contribute in each year and the number of years that they then go on to
remain active in the mailing lists. For example, a participant who first
sent an e-mail to an IETF list in 2010, and last sent an e-mail to an IETF
list in 2018, will have a contribution duration of 9 years.

We generate Gaussian Mixture Models for observing different clusters of the
maximum duration that contributors remain active for.  These models
identified three broad clusters that contributors can be categorised
within:

\begin{itemize}
    \item \emph{Young contributors} who leave within 1 year of joining.
    \item \emph{Mid-age contributors} who go onto remain associated for
      more than 1 year, but less than 5 years.
    \item \emph{Senior contributors} who go onto remain associated for 5 or
      more years.
\end{itemize}

We use these contribution categories defined above to characterise the
interactions between contributors belonging to different categories, as
well as the authors of each RFC. We look at outgoing and incoming
interactions with RFC authors in the period between the first draft and
publication of the RFC. If this period is less than two years, we look at
the activities of authors for two years before the RFC was published.
Interactions are defined from the viewpoint of the RFC authors:

\begin{itemize}
    \item \textit{Outgoing interaction}, where an RFC author responds to an
      email from other contributors (i.e., email sent).
    \item \textit{Incoming interaction}, where contributors respond to an
      email by the author (i.e., email received).
\end{itemize}

For each interaction type, we count the number of messages and number of
unique contributors involved in those interactions. For each RFC, we
analyse the interaction activities based on three measures of author
contribution duration:

\begin{itemize}
    \item \emph{Junior-most author}, i.e., the author with the lowest
      contribution duration at the time of publication.
    \item \emph{Senior-most author}, i.e., the author with the highest
      contribution duration at time of publication.
    \item \emph{Mean contribution duration} of all of the authors of the
      RFC at the time of publication.
\end{itemize}

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/age_authors_RFCs.pdf}
\caption{Contribution duration distribution of authors of RFCs: junior-most
  author of each RFC, senior-most author of each RFC, and mean
  contribution-duration of all authors of each RFC}
\label{fig:age_dist_rfc_authors}
\end{figure}

Figure~\ref{fig:age_dist_rfc_authors} shows the distribution of
contribution duration of each of these three measures. This shows that the
majority of junior-most authors have participated for less than 5 years in
the IETF, whereas the majority of senior-most authors have participated in
the IETF for more than 10 years. In fact, 35\% of authors exceed 15 years
of IETF participation, showing that RFCs tend to be authored by a mix of
seniority levels.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/rfc_authors_degree.pdf}
\caption{CDF showing drift in annual degree (interaction with their
  network) of RFC authors over the period 2000-2020.}
\label{fig:degree_dist_rfc_authors}
\end{figure}

%..................................................................................................
\pb{Evolution of interactions:}
Figure~\ref{fig:degree_dist_rfc_authors} shows the gradual drift in annual
degree (i.e., the number of people interacted with) of RFC authors. The
degree of authors has substantially increased over time. For instance, in
the year 2000 only 5.5\% of the authors had a degree of over 25, whereas by
the year 2015 almost a quarter of the authors had a degree over 25.  This
confirms that, on average, more recent RFCs generate greater discussion.
This may explain increasing publication times: authors spend more time
interacting with other participants, and these discussions are likely to
lead to more drafts.
\todo{Replace with material from one of the later papers?}

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures-prev/imc-2021/emails/junior_senior_authors_seniormem_indegree.pdf}
\caption{CDF showing number of senior contributors sending messages
  (in-degree) to junior and senior authors.}
\label{fig:junior_senior_indegree_seniormem}
\end{figure}

We can also contrast the interaction patterns of the junior vs.\ senior
authors. Figure~\ref{fig:junior_senior_indegree_seniormem} presents the
CDFs of the in-degree across both junior and senior authors.  It shows that
the incoming interactions from senior contributors to junior authors are
significantly less than the incoming interactions from senior contributors
to senior authors. Nearly 55\% of junior authors receive messages from
fewer than 10 senior contributors, whereas nearly 65\% of senior authors
receive messages from more than 10 senior contributors.  This speaks to the
differing roles played by these sub-populations: senior authors act as hubs
through which substantial volumes of interaction flow.
\todo{Replace with material from one of the later papers?}

%..................................................................................................
\pb{Summary:}
Emails play a vital role in underpinning RFC publication with approximately
130,000 emails sent per year. We have shown that the seniority of both
participants and RFC authors fundamentally changes the volume of
interactions that they have. These trends are likely to have implications
for the IETF community, especially as it tries to encourage new
participants.

%--------------------------------------------------------------------------------------------------
\subsection{Influence}

% ICWSM 2022 paper
% - Measuring influence
% - Behaviour of influential participants
% - Impact of influential participants

%--------------------------------------------------------------------------------------------------
\subsection{Hierarchy}
%..................................................................................................
% ICWSM 2024 paper
% - RQ1: Are higher levels of the hierarchy growing or becoming more
%   centralised? Are they associated with an increased domination of the
%   conversation? 
% - RQ2: What is the association between the organisation hierarchy and
%   general communication patterns?
% - RQ3: How do people with differing roles communicate with each other?
%   Does information flow “up” or “down” the hierarchy? 
% - RQ4: What is the impact that individuals have on their direct contacts’
%   activity over time? How does this vary across hierarchy level?

%..................................................................................................
% LEDA
% - a taxonomy of dialogue acts (DAs) and a labeled dataset of emails

%==================================================================================================
\section{Understanding Participant Behaviour}

% Tracing Linguistic Markers
% - RQ1: How do linguistic traits differ between more and less influential
%   participants? 
% - RQ2: How do linguistic traits vary for participants at different levels
%   of the organisation hierarchy? 
% - RQ3: How does linguistic behaviour of participants change as they gain
%   influence?

% Frontiers in Psychology paper
% - Use of Language, discussing how it shifts as people become more senior

%==================================================================================================
\section{Success Factors}

\subsection{For authors}
% ICWSM 2022 paper Section 4

\subsection{For documents}

% IMC 2021 paper Section 4
